---
title: "Endogenous DNA fraction of sirenians from historical collections"
output: html_notebook 'Endogenous DNA fraction'
---

Introduction: In this paper we investigate aDNA sequencing success in a dataset 
of large, tropical marine mammals –  sirenians – sampled at a variety of 
European NHCs. We consider 91 samples from 85 specimens comprising five species. 
We compare sample types, species and region, individual collections, and 
material age, to assess what contributes most towards the successful sequencing 
of specimens. Additionally, we explore the potential of soft tissue and more 
specifically dried cartilage for aDNA research. We make considerations of how 
these data can inform the aDNA workflow to minimise destruction of natural 
history and maximise DNA yield.

Note: Hydrodamalis gigas (aka Steller's sea cow) is referred to in this script
as SSC

Load the packages needed:
```{r}
library(readxl)
library(dplyr)
library(car)
library(lme4)
library(ggplot2)
library(ggpubr)
```

Read the data from the excel sheet called Tables_Methods_NEW using readxl

################################################################################
################################################################################
################################################################################

COUNTS AND DATA EXPLORATION

First lets get an idea of what the data look like and get some data counts for 
our predictor variables of interest
```{r}
summary(data)
str(data)
# Summary for variables
char_variables <- c("Species", "Sample type", "Sample type class"
                    , "Anatomy class", "Institution code", "Region", 
                     "B/S 1950", "Year (bins) new") 
summary_list <- lapply(data[, char_variables], table)
print(summary_list)

# Summary of sequencing
sum(data$`Pair(n)`)
# we obtained ~6.25 billion paired total sequencing reads
mean(data$`Length(bp)`)
# mean read length of 79.1bp
range(data$`Endogenous(fraction)`)
# 0.0002-69% endogenous DNA
range(data$Clonality)
# 0.02-21% clonality
range(data$Nu_coverage)
# nuclear coverage 0-11.32-fold
count_table <- data %>%
  filter(`Endogenous(fraction)` >= 0.2)
(nrow(count_table)/nrow(data))*100
# ~56% of samples have >20% endogenous

# Now i'll subset the dataframe to only keep the columns we are interested in
data <- data[, c("aDNA lab ID", "Endogenous(fraction)", 
                 "Species", "Sample type", "Sample type class", 
                 "Anatomy class", "Institution code", "Region", 
                 "B/S 1950", "Year (bins) new")]
```
This information is given in Table 1 and Figures 1 and 2 in the manuscript

There were two individuals that could not confidently be assigned to Sirenia
Remove them...
```{r}
data <- data %>% filter(`aDNA lab ID` != "STM004" & `aDNA lab ID` != "STM017")
```

VISUALISATION OF DATA
Use violin and dot plots to visualise endogenous DNA fraction across the
predictor variables that we are interested in

```{r}
# I'll remove all NAs for the anatomy class plot (this is the soft tissue)
Anatomyplotting <-data %>% filter(`Anatomy class` != "NA")
```

Plot...
```{r}
############################ SPECIES ############################
# Set the desired order...
data$`Species` <- factor(data$`Species`, 
                                 levels = c("Hydrodamalis gigas", "Dugong dugon", "Trichechus manatus", "Trichechus inunguis", "Trichechus senegalensis"))

# First get the counts for n
# AND calculate the number of points and maximum value for each 
species_stats <- data %>% 
  group_by(`Species`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Species_violin <- ggplot(data, aes(x=Species, y=`Endogenous(fraction)`, fill = Species)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("skyblue",5)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("Species") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Hydrodamalis gigas","Dugong dugon", "Trichechus manatus", "Trichechus inunguis", "Trichechus senegalensis")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = species_stats, aes(x = Species, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Species_violin

############################ SAMPLE TYPE ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
sampletype_stats <- data %>% 
  group_by(`Sample type`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Sampletype_violin <- ggplot(data, aes(x=`Sample type`, y=`Endogenous(fraction)`, fill = `Sample type`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("pink",2)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("Sample type") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Bone", "Soft tissue")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = sampletype_stats, aes(x = `Sample type`, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Sampletype_violin

############################ SAMPLE TYPE CLASS ############################
# Set the desired order...
data$`Sample type class` <- factor(data$`Sample type class`, 
                                 levels = c("Cranial", "Post-cranial", 
                                            "Cartilage", "Skin/Hide", "Unknown"))
# Now get the counts for n
# AND calculate the number of points and maximum value for each 
sampletypeclass_stats <- data %>% 
  group_by(`Sample type class`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Sampletypeclass_violin <- ggplot(data, aes(x=`Sample type class`, y=`Endogenous(fraction)`, fill = `Sample type class`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("palegreen",5)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("Sample type class") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Cranial", "Post-cranial", "Cartilage", "Skin/Hide", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = sampletypeclass_stats, aes(x = `Sample type class`, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Sampletypeclass_violin

############################ ANATOMY CLASS ############################
# We need to first remove NA (this is the soft tissue samples)
Anatomyplotting <- data %>% filter(`Anatomy class` != "NA")

# Let's order by most to least mean endogenous
Anatomyplotting$`Anatomy class` <- factor(Anatomyplotting$`Anatomy class`)
mean_endogenous <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  summarise(mean_endogenous = mean(`Endogenous(fraction)`)) %>%
  arrange(desc(mean_endogenous))
# reorder AND place the "Unknown" category last
ordered_levels <- mean_endogenous$`Anatomy class`
if("Unknown" %in% ordered_levels) {
  ordered_levels <- c(setdiff(ordered_levels, "Unknown"), "Unknown")
}
Anatomyplotting$`Anatomy class` <- factor(Anatomyplotting$`Anatomy class`, levels = ordered_levels)

# Now get the counts for n and calculate the number of points and maximum value for each
anatomyclass_stats <- Anatomyplotting %>% 
  group_by(`Anatomy class`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Divide data based on the number of points (< 3 or => 3)
data_few_points <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  filter(n() < 3)
data_many_points <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  filter(n() >= 3)

# Create the plot using dotplot of entire dataset and then overlay violin plots of the 'many points' data
Anatomyclass_violin <- ggplot() +
  geom_dotplot(data = Anatomyplotting, aes(x = `Anatomy class`, y = `Endogenous(fraction)`, fill = `Anatomy class`), binaxis = "y", stackdir = "center", dotsize = 40, binwidth = 0.0003) +
  geom_violin(data = data_many_points, aes(x = `Anatomy class`, y = `Endogenous(fraction)`, fill = `Anatomy class`), alpha = 0.6, lwd = 0.3) +
  scale_fill_manual(name = "Endogenous(fraction)", values = rep("orange", 19)) + 
  xlab("Anatomy class") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels = ordered_levels) +
  scale_y_continuous(breaks = seq(0.0, 1.0, by = 0.1), limits = c(0.0, 1.0)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size = 11, angle = 45, vjust = 1, hjust = 1)) +
  geom_text(data = anatomyclass_stats, aes(x = `Anatomy class`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Anatomyclass_violin

############################ INSTUTUTION CODE ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
institution_stats <- data %>% 
  group_by(`Institution code`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))

# We have some categories with too few points create a violin plot so need to divide the data
# Separate data based on the number of points
data_few_points <- data %>%
  group_by(`Institution code`) %>%
  filter(n() < 3)

data_many_points <- data %>%
  group_by(`Institution code`) %>%
  filter(n() >= 3)

# Create the plot
Institution_violin <- ggplot() +
  geom_violin(data = data_many_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), alpha = 0.6, lwd = 0.3) + 
  geom_dotplot(data = data_many_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), binaxis = "y", stackdir = "center", dotsize = 40, binwidth = 0.0003) + 
  geom_dotplot(data = data_few_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), binaxis = "y", stackdir = "center", dotsize = 40, binwidth = 0.0003) +
  scale_fill_manual(name = "Endogenous(fraction)", values = rep("red", 7)) + 
  xlab("Institution code") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels = c("MCC", "MNHN", "NBC", "NHMO", "NHMUK", "RMCA", "UMB")) +
  scale_y_continuous(breaks = seq(0.0, 1.0, by = 0.1), limits = c(0.0, 1.0)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size = 11, angle = 45, vjust = 1, hjust = 1)) +
  geom_text(data = institution_stats, aes(x = `Institution code`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Institution_violin

############################ REGION ############################
# First, calculate counts for n and max values for each region
region_stats <- data %>% 
  group_by(`Region`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))

# Set the order for the regions, with "Unknown" as the last category
region_order <- c("Amazon", "Caribbean and Gulf of Mexico", "Central Africa", 
                  "Eastern Indian Ocean", "North Pacific", "South America", 
                  "West Africa", "Western Indian Ocean", "Unknown")
data$Region <- factor(data$Region, levels = region_order)

# Create the plot
Region_violin <- ggplot(data, aes(x = Region, y = `Endogenous(fraction)`, fill = Region)) +
  geom_violin(alpha = 0.6, lwd = 0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)", values = rep("yellow", length(region_order))) +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 40, binwidth = 0.0003) +
  xlab("Region") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels = region_order) +
  scale_y_continuous(breaks = seq(0.0, 1.0, by = 0.1), limits = c(0.0, 1.0)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size = 11, angle = 45, vjust = 1, hjust = 1)) +
  geom_text(data = region_stats, aes(x = `Region`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Region_violin

############################ YEARS B/S 1950 ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
years_stats <- data %>% 
  group_by(`B/S 1950`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
Years_violin <- ggplot(data, aes(x=`B/S 1950`, y=`Endogenous(fraction)`, fill = `B/S 1950`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("darkgreen",5)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("B/S 1950") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Before","Since", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = years_stats, aes(x = `B/S 1950`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Years_violin

### Let's try replotting this with the increased detail given in the Year (bins) new column...

############################ YEARS (binned) NEW ############################
# Set the desired order...
data$`Year (bins) new` <- factor(data$`Year (bins) new`, 
                                 levels = c("Pre-1800", "1800-1899", "1900-1949", 
                                            "1950-1999", "Pre-1900 Unknown", 
                                            "Pre-1950 Unknown", "Unknown"))
# Now get the counts for n
# AND calculate the number of points and maximum value for each 
years_stats2 <- data %>% 
  group_by(`Year (bins) new`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
Years_violin2 <- ggplot(data, aes(x=`Year (bins) new`, y=`Endogenous(fraction)`, fill = `Year (bins) new`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("darkgreen",7)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("Year (bins) new") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Pre-1800", "1800-1899", "1900-1949", 
                            "1950-1999", "Pre-1900 Unknown", 
                            "Pre-1950 Unknown", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = years_stats2, aes(x = `Year (bins) new`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Years_violin2
```

Interpretation:
At first glance it looks like endogenous fraction is quite variable across 
predictors - the only exception seems to be with temporal predictors because 
there are few differences across binned time periods that are clear to 
understand.
Clearly Pre-1800 SSC (Hydrodamalis gigas) is worse than other binned time 
periods (this is the only obvious pattern) but all 15 samples in this category 
come from one species (which is not represented in other time periods)
from a specific region and unique depositional environment...

Test to see if there are significant differences between the time periods
```{r}
# Do t.test for Before vs Since 1950
x <- subset(data,`B/S 1950` == "Before") 
y <- subset(data,`B/S 1950` == "Since")
t.test(x$`Endogenous(fraction)`,y$`Endogenous(fraction)`,var.equal = FALSE)
# No significant difference - p-value = 0.9546 - means are almost identical

# Do ANOVA for binned periods
one.way.aov1 <- aov(`Endogenous(fraction)`~`Year (bins) new`, data = data)
summary(one.way.aov1)
# Filter all unknown categories out...
filter_years_data1 <- data %>% 
  filter(!`Year (bins) new` %in% c("Unknown", "Pre-1900 Unknown", "Pre-1950 Unknown"))
one.way.aov2 <- aov(`Endogenous(fraction)`~`Year (bins) new`, data = filter_years_data1)
summary(one.way.aov2)
# Filter SSC category out...
filter_years_data2 <- filter_years_data1 %>% 
  filter(!`Year (bins) new` %in% c("Pre-1800 SSC"))
one.way.aov3 <- aov(`Endogenous(fraction)`~`Year (bins) new`, data = filter_years_data2)
summary(one.way.aov3)
```
Conclusion: The means of each binned subset are significantly different when 
using 1) all bins 2) all bins except for the 3 with uncertainty of date, BUT in 
3) once the Pre-1800 SSC (Hydrodamalis gigas) bin is removed there is no longer 
a significant difference across time periods...

Test to see if there are significant differences between the other variables
```{r}
### Species
one.way.aov4 <- aov(`Endogenous(fraction)`~`Species`, data = data)
summary(one.way.aov4)
# Test without Hydrodamalis as this appears to be uniquely bad based on the plot
filter_sp_data <- data %>% 
  filter(!`Species` %in% c("Hydrodamalis gigas"))
one.way.aov5 <- aov(`Endogenous(fraction)`~`Species`, data = filter_sp_data)
summary(one.way.aov5)

### Sample type class
# Filter unknown categories out...
filter_stc_data <- data %>% 
  filter(!`Sample type class` %in% c("Unknown"))
one.way.aov6 <- aov(`Endogenous(fraction)`~`Sample type class`, data = filter_stc_data)
summary(one.way.aov6)

### Institution
one.way.aov7 <- aov(`Endogenous(fraction)`~`Institution code`, data = data)
summary(one.way.aov7)
### Region
# Filter unknown out...
filter_region_data1 <- data %>% 
  filter(!`Region` %in% c("Unknown"))
one.way.aov8 <- aov(`Endogenous(fraction)`~`Region`, data = filter_region_data1)
summary(one.way.aov8)
# Now filter the North Pacific out as this is only made up from the SSC which were shown in the Species ANOVA to affect the result...
filter_region_data2 <- filter_region_data1 %>% 
  filter(!`Region` %in% c("North Pacific"))
one.way.aov9 <- aov(`Endogenous(fraction)`~`Region`, data = filter_region_data2)
summary(one.way.aov9)
```

Conclusions: 
1) There is a significant difference between species BUT only when you include the SSCs - 
   If you remove them then you see that there is no significant difference between extant sirenian species.
2) The ANOVA for sample type class demonstrates that there is a significant difference between sample types.
   (See below section for t.test of bone/soft tissue)
   This, coupled with the anatomy plot which shows very variable endogenous DNA across all elements
   sampled, demonstrates that we need to investigate these differences further... 
3) There is no significant difference between institutions
4) Region also becomes not significant once the unique North Pacific (SSC) samples are removed (p=0.635)

Clearly the SSC material is unique(ly bad) and i don't want it to impact my interpretation
of the data or understanding of what it actually is that leads to the successful sequencing
of historical and archaeological NHC material. This is already an interesting finding in-
and-of-itself as we usually expect cooler latitudes to yield better DNA preservation than 
tropical and temperate latitudes. Clearly there is something about the depositional environment
and/or the subsequent unrecorded treatment of the bones between collection
and our sampling that made these bones particularly bad for sequencing.
For this study we will now continue without these SSC data...

I have rewritten all of the plotting and these tests with none of the SSC
data - it's in the script Supp_Script_v2.Rmd and given in the main manuscript text

This script continues from now without any of the SSC data.

```{r}
# Remove SSCs 
data2 <- data %>% 
  filter(!`Species` %in% c("Hydrodamalis gigas"))
summary(data2)
mean(data2$`Endogenous(fraction)`)
```
Mean endogenous for data2 is 33.8%

################################################################################
################################################################################
################################################################################
EXPLORATION OF SAMPLE TYPES...
USE T TESTS TO ASSESS FOR SIGNIFICANT DIFFERENCES BETWEEN DEFINED SUBSETS...

Now lets use a Student's test using the t.test R function to assess if there 
is a significant difference between endogenous DNA fraction in different sample type subsets
that we can create from our dataset
Note: Our sample sizes are not equal so we will use var.equal=FALSE

CRANIAL VS POSTCRANIAL BONE
```{r}
cranial <- data2 %>% filter(`Sample type class` == "Cranial")
postcranial <- data2 %>% filter(`Sample type class` == "Post-cranial")

t.test(cranial$`Endogenous(fraction)`,postcranial$`Endogenous(fraction)`
       , var.equal = FALSE)
```
Cranial and postcranial bone are significantly different based on t.test

IS PETROUS BONE BETTER THAN TYMPANIC?
```{r}
petrous <- data2 %>% filter(`Anatomy class` == "Petrous")
tympanic <- data2 %>% filter(`Anatomy class` == "Tympanic")

t.test(petrous$`Endogenous(fraction)`,tympanic$`Endogenous(fraction)`
       , var.equal = FALSE)
```
Petrous and tympanic are not significantly different based on t.test

IS PETROUS BONE BETTER THAN ALL OTHER TYPES OF BONE?
```{r}
all_other <- data2 %>% filter(`Sample type` == "Bone")
# remove the ambiguous PT category so as not to confound
all_other <- all_other %>% filter(`Anatomy class` != "Petrous" & 
                                    `Anatomy class` != "PT")

t.test(petrous$`Endogenous(fraction)`,all_other$`Endogenous(fraction)`
       , var.equal = FALSE)
```
Petrous bone is not significantly better based on t.test

WHAT ABOUT PETROUS COMPARED ONLY TO THE REST OF THE CRANIUM?
```{r}
# remove ambiguous PT category so as not to confound
cranial_wout_petrous <- cranial %>% filter(`Anatomy class` != "Petrous" & 
                                             `Anatomy class` != "PT")
t.test(petrous$`Endogenous(fraction)`,cranial_wout_petrous$
         `Endogenous(fraction)`, var.equal = FALSE)
```
Petrous bone is not significantly better when only comparing within the cranium 

WHAT ABOUT PETROUS AND TYMPANIC COMPARED TO THE REST OF THE CRANIUM...?
```{r}
cranial_wout_PT <- cranial %>% filter(`Anatomy class` != "Petrous" & 
                                        `Anatomy class` != "Tympanic" & 
                                        `Anatomy class` != "PT")
PT <- cranial %>% filter(`Anatomy class` == "Petrous" | 
                           `Anatomy class` == "Tympanic" | 
                           `Anatomy class` == "PT")
t.test(PT$`Endogenous(fraction)`,cranial_wout_PT$`Endogenous(fraction)`
       , var.equal = FALSE)
```
Highly significantly different!

...AND TO THE REST OF THE SKELETON?
```{r}
skeleton_wout_PT <- data %>% filter(`Sample type` == "Bone")
skeleton_wout_PT <- data %>% filter(`Anatomy class` != "Petrous" & 
                                        `Anatomy class` != "Tympanic" & 
                                        `Anatomy class` != "PT")
t.test(PT$`Endogenous(fraction)`,skeleton_wout_PT$`Endogenous(fraction)`
       , var.equal = FALSE)
```
Even more highly significantly different!

LET'S JUST SEE WHAT WE GET IF WE TRY THIS WITH TYMPANIC INSTEAD...
```{r}
all_other2 <- data2 %>% filter(`Sample type` == "Bone")
# remove ambiguous PT category so as not to confound
all_other2 <- all_other2 %>% filter(`Anatomy class` != "Tympanic" & 
                                    `Anatomy class` != "PT")
# to the entire rest of the skeleton...
t.test(tympanic$`Endogenous(fraction)`,all_other$`Endogenous(fraction)`
       , var.equal = FALSE)
# just within the cranium...
cranial_wout_tympanic <- cranial %>% filter(`Anatomy class` != "Tympanic" & 
                                             `Anatomy class` != "PT")
t.test(tympanic$`Endogenous(fraction)`,cranial_wout_tympanic$
         `Endogenous(fraction)`, var.equal = FALSE)
```
No significant difference...

What's going on here clearly requires further investigation to be untangled further
... It's the PT category unfortunately driving this significant difference

HOW DO BONE AND SOFT TISSUE COMPARE?
```{r}
# Bone versus soft tissue
bone <- data2 %>% filter(`Sample type` == "Bone")
soft_tissue <- data2 %>% filter(`Sample type` == "Soft tissue")
t.test(bone$`Endogenous(fraction)`,soft_tissue$`Endogenous(fraction)`
       , var.equal = FALSE)
# Bone versus cartilage only
cartilage <- data2 %>% filter(`Sample type class` == "Cartilage")
t.test(bone$`Endogenous(fraction)`,cartilage$`Endogenous(fraction)`
       , var.equal = FALSE)

```
No significant differences

Out of interest...because the distribution for soft tissues is noticably bimodal...
HOW DO THE TWO HIGH YIELD SOFT TISSUE SAMPLES IMPACT OUR TESTS?
```{r}
soft_tissue_modified <- soft_tissue %>% filter(`aDNA lab ID` != "STM018" &
                                                 `aDNA lab ID` != "SAF023")
mean(soft_tissue_modified$`Endogenous(fraction)`)
t.test(bone$`Endogenous(fraction)`,soft_tissue_modified$`Endogenous(fraction)`
       , var.equal = FALSE)
```
When they are removed there is a significant difference - so soft tissue seems to 
be quite highly variable in endogenous DNA yield... without these two potentially
exceptional samples the difference in means is 10-fold.

ANOVA of anatomy
```{r}
Anatomy_aov <-Anatomyplotting %>% filter(`Anatomy class` != "Unknown")
Anatomy_aov <-Anatomy_aov %>% filter(`Species` != "Hydrodamalis gigas")
one.way.aov10 <- aov(`Endogenous(fraction)`~`Anatomy class`, data = Anatomy_aov)
summary(one.way.aov10)
```

Interpretation:
So far it looks like 1) Cranial bone has greater sequencing success than post-
cranial bone, and 2) Petrous AND tympanic together show a significant difference in 
sequencing success when compared to other subsets - but further exploration is 
needed to tease apart any significant differences within the earbone complex as 
individually we found no significant differences...
4) there is not a significant difference when comparing bone and soft tissue...
5) I think most of the variation is coming from within the anatomy types but this is 
difficult to fully understand based on our sample size

################################################################################
################################################################################
################################################################################

USING A LINEAR REGRESSION MODEL, ASSESS WHICH VARIABLES ARE MOST INFLUENTIAL
IN DETERMINING SEQUENCING SUCCESS

```{r}
# Convert "Unknown" to NA and "NA" text to R readable NA, then...
# convert all categorical predictor variables to 
# factors and check data structure
data3 <- data2 %>%
  mutate(across(c(Species, `Sample type`, `Sample type class`, `Anatomy class`, 
                  `Institution code`, `Region`, `B/S 1950`, `Year (bins) new`), 
                ~ na_if(., "Unknown"))) %>%
  mutate(across(where(is.character), ~ na_if(., "NA")),
         Species = factor(Species),
         `Sample type` = factor(`Sample type`),
         `Sample type class` = factor(`Sample type class`),
         `Anatomy class` = factor(`Anatomy class`),
         `Institution code` = factor(`Institution code`),
         `Region` = factor(`Region`),
         `B/S 1950` = factor(`B/S 1950`),
         `Year (bins) new` = factor(`Year (bins) new`))

summary(data3)

# Cross check to confirm correct columns have NA
sapply(data3, function(x) any(x == "NA"))
```

Let's see how endogenous fraction is varying across the skeleton and
between sample types...
```{r}
### SAMPLE TYPE
anatomymodel1 <- lm(data = data3, `Endogenous(fraction)` ~ `Sample type`)
summary(anatomymodel1)
# the model is significant (p = 0.02567) but on the whole only explains around
# 5% of the variation in endogenous fraction
# Soft tissue samples have an estimated -0.19213 lower mean compared to bone

### SAMPLE TYPE CLASS
anatomymodel2 <- lm(data = data3, `Endogenous(fraction)` ~ `Sample type class`)
summary(anatomymodel2)
# Cranial has a statistically significant positive effect (p = 0.0194)
# the model overall is significant (p-value: 0.01761) and has greater explanatory
# power than the previous model (~9% of variation explained by this predictor) 
# appears to be a better overall fit for the data - the more levels in this 
# model provide a more nuanced analysis - the model also has lower residual 
# standard error

### ANATOMY CLASS
# now lets look at anatomy more specifically...
# this was difficult to gauge from the plot as we have so many different 
# categories
anatomymodel3 <- lm(data = data3, `Endogenous(fraction)` ~ `Anatomy class`)
summary(anatomymodel3)
# in this model the spread in the residuals (from -0.39768 to 0.27132) is 
# tighter than in the previous models, suggesting that the model predictions 
# align closely with the observed data
# PT is a significantly positive predictor (p value = 0.0161) and petrous alone 
# is marginally significant (>0.1)
# the model is statistically significant (p-value: 0.02541) and explains far
# more of the variation in endogenous fraction (Adjusted R-squared:  0.2571) 
# than the previous models (although some explanatory power is lost due to high
# number of predictor levels - Multiple R-squared:  0.4772) - therefore although
# the model is significant many predictor levels are not contributing significantly
# increased complexity may be detrimental - 21 observations were also deleted 
# due to missingness
```

For fine scale effects the third model may provide more insights and greater detail
but for overall interpretability the second is likely better for our analysis

Check the other predictor variables INDIVIDUALLY...
```{r}
species_model <- lm(data = data3, `Endogenous(fraction)` ~ `Species`)
summary(species_model)
# model does nothing for explaining the variation in endogenous fraction
# F-statistic: 0.7409 on 3 and 72 DF,  p-value: 0.5311

institution_model <- lm(data = data3, `Endogenous(fraction)` ~ `Institution code`)
summary(institution_model)
# The model explains around 5% of the variation observed
# Overall it is not significant (F-statistic: 1.789 on 6 and 69 DF,  p-value: 0.1141)
# however, significance of individual levels NHMO + NHMUK (P VALUE >0.05) 
# suggests that, although the model itself is not significant, there is 
# significant variation between levels that we might explore further...

region_model <- lm(data = data3, `Endogenous(fraction)` ~ `Region`)
summary(region_model)
# model does nothing for explaining the variation in endogenous fraction
# F-statistic: 0.7198 on 6 and 64 DF,  p-value: 0.6351

time_model_1 <- lm(data = data3, `Endogenous(fraction)` ~ `B/S 1950`)
summary(time_model_1)
# there is a slight but non-significant decrease between before and since 1950
# and overall the model does little for explaining the variation in endogenous 
# fraction - F-statistic: 1.832 on 1 and 52 DF,  p-value: 0.1818

time_model_2 <- lm(data = data3, `Endogenous(fraction)` ~ `Year (bins) new`)
summary(time_model_2)
# F-statistic: 2.431 on 4 and 49 DF,  p-value: 0.05992
# As above model...
```

TEST FIT A MODEL WITH MULTIPLE PREDICTORS
Fit the model with the main predictors that seem so far like they 
could be impactful on the success of sequencing
```{r}
model <- lm(data = data3, `Endogenous(fraction)` ~ `Sample type class` + 
              `Anatomy class` + `Institution code`)
summary(model)
drop1(model)

# USING CAR PACKAGE - vif() function - TO MEASURE VIFs
vif(model)
# cannot calculate due to aliased coefficients
alias(model)
# there are many!
```

Interpretation: Can't use all 3 predictors (unsurprisingly!) due to aliased 
coefficients - losing too many data points...

I'll try simplifying the model and measuring collinearity...
```{r}
model2 <- lm(data = data3, `Endogenous(fraction)` ~ `Sample type class` + 
               `Institution code`)
summary(model2)
vif(model2)

model3 <- lm(data = data3, `Endogenous(fraction)` ~ `Anatomy class` + 
               `Institution code`)
summary(model3)
vif(model3)
```

Intrpretation:
The first model explains far less of the variance - adjusted R2: 14.5% vs 51% 
in endogenous DNA fraction. The second model has a lower residual standard error
indicating a better fit to the data. Both are statistically significant:
Model 2: F-statistic: 2.375 on 9 and 64 DF,  p-value: 0.02188
Model 3: F-statistic: 3.686 on 21 and 33 DF,  p-value: 0.000406
... model 3 is however, very highly statistically significant.
All GVIF values are below 2 which suggests that there is no significant 
multicollinearity. 
HOWEVER: The 21 deleted observations in model 3 may be introducing bias into the
dataset - the differences in individual levels for institution seem very 
different when comparing the two models which is suspicious.

Look at model 2 again:
```{r}
summary(model2)
vif(model2)
anova(model2)
drop1(model2, test = "Chisq")
```
Single term deletions:
Although Institution might be a predictor of 
interest, it is only removal of Sample type class that will significantly 
worsen the fit of the model (F = 3.8042, p-value = 0.01421).

Akaike Information Criterion (AIC) and RSS:
The full model, including both predictors has an AIC of -210.53 and RSS of 3.2833.
With Sample type class removed, RSS increased (3.7707) and AIC worsened to -206.28, 
a highly significant (p-value = 0.01662) change. By contrast, with Institution 
code removed RSS increased very little (3.7945) and AIC increased slightly
to -211.82 (p-value = 0.09783). In all this suggests that Sample type class is the 
significant predictor in the model, and that Institution code can be removed to simplify 
the model.
