---
title: "Endogenous DNA fraction of sirenians from historical collections"
output: html_notebook 'Endogenous DNA fraction'
---

### This version of the plotting script creates the plots and runs the tests
### without the extinct species - Hydrodamalis gigas - included in the dataset

Load the packages needed:
```{r}
library(readxl)
library(dplyr)
library(car)
library(lme4)
library(ggplot2)
library(ggpubr)
```

Read the data from the excel sheet called Tables_Methods_NEW using readxl

################################################################################
################################################################################
################################################################################

There were two individuals that could not confidently be assigned to Sirenia
They are removed from the study...
```{r}
data <- data %>% filter(`aDNA lab ID` != "STM004" & `aDNA lab ID` != "STM017")
```

VISUALISATION OF DATA
Use violin and dot plots to visualise endogenous DNA fraction across the
predictor variables that we are interested in

```{r}
# Remove SSCs 
data2 <- data %>% 
  filter(!`Species` %in% c("Hydrodamalis gigas"))
summary(data2)
mean(data2$`Endogenous(fraction)`)
```
Mean endogenous DNA fraction is greater (~34%) in this dataset

Plot...
```{r}
############################ SPECIES ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
species_stats <- data2 %>% 
  group_by(`Species`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Species_violin <- ggplot(data2, aes(x=Species, y=`Endogenous(fraction)`, fill = Species)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("skyblue",4)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 40,
               binwidth = 0.0003) +
  xlab("Species") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Dugong dugon","Trichechus inunguis", "Trichechus manatus", "Trichechus senegalensis")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.1),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 5)),
        axis.text.x = element_text(size=11, angle = 45, vjust = 1, hjust=1)) +
  geom_text(data = species_stats, aes(x = Species, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Species_violin

############################ SAMPLE TYPE ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
sampletype_stats <- data2 %>% 
  group_by(`Sample type`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Sampletype_violin <- ggplot(data2, aes(x=`Sample type`, y=`Endogenous(fraction)`, fill = `Sample type`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("pink",2)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 20,
               binwidth = 0.0001) +
  xlab("Sample type") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Bone", "Soft tissue")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.2),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 5),
        axis.text.x = element_text(size=5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size=5)) +
  geom_text(data = sampletype_stats, aes(x = `Sample type`, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Sampletype_violin

############################ SAMPLE TYPE CLASS ############################
# Set the desired order...
data2$`Sample type class` <- factor(data2$`Sample type class`, 
                                 levels = c("Cranial", "Post-cranial", 
                                            "Cartilage", "Skin/Hide", "Unknown"))
# Now get the counts for n
# AND calculate the number of points and maximum value for each 
sampletypeclass_stats <- data2 %>% 
  group_by(`Sample type class`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Then make the plot...
Sampletypeclass_violin <- ggplot(data2, aes(x=`Sample type class`, y=`Endogenous(fraction)`, fill = `Sample type class`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("palegreen",5)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 20,
               binwidth = 0.0001) +
  xlab("Sample type class") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Cranial", "Post-cranial", "Cartilage", "Skin/Hide", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.2),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 5),
        axis.text.x = element_text(size=5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size=5)) +
  geom_text(data = sampletypeclass_stats, aes(x = `Sample type class`, y = max_y + 0.1, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Sampletypeclass_violin

############################ ANATOMY CLASS ############################
# We need to first remove NA (this is the soft tissue samples)
Anatomyplotting <-data2 %>% filter(`Anatomy class` != "NA")

# Let's order by most to least mean endogenous
Anatomyplotting$`Anatomy class` <- factor(Anatomyplotting$`Anatomy class`)
mean_endogenous <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  summarise(mean_endogenous = mean(`Endogenous(fraction)`)) %>%
  arrange(desc(mean_endogenous))
# reorder AND place the "Unknown" category last
ordered_levels <- mean_endogenous$`Anatomy class`
if("Unknown" %in% ordered_levels) {
  ordered_levels <- c(setdiff(ordered_levels, "Unknown"), "Unknown")
}
Anatomyplotting$`Anatomy class` <- factor(Anatomyplotting$`Anatomy class`, levels = ordered_levels)

# Now get the counts for n and calculate the number of points and maximum value for each
anatomyclass_stats <- Anatomyplotting %>% 
  group_by(`Anatomy class`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
# Divide data based on the number of points (< 3 or => 3)
data_few_points <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  filter(n() < 3)
data_many_points <- Anatomyplotting %>%
  group_by(`Anatomy class`) %>%
  filter(n() >= 3)

# Create the plot using dotplot of entire dataset and then overlay violin plots of the 'many points' data
Anatomyclass_violin <- ggplot() +
  geom_dotplot(data = Anatomyplotting, aes(x = `Anatomy class`, y = `Endogenous(fraction)`, fill = `Anatomy class`), binaxis = "y", stackdir = "center", dotsize = 40, binwidth = 0.0003) +
  geom_violin(data = data_many_points, aes(x = `Anatomy class`, y = `Endogenous(fraction)`, fill = `Anatomy class`), alpha = 0.6, lwd = 0.3) +
  scale_fill_manual(name = "Endogenous(fraction)", values = rep("orange", 19)) + 
  xlab("Anatomy class") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels = ordered_levels) +
  scale_y_continuous(breaks = seq(0.0, 1.0, by = 0.2), limits = c(0.0, 1.0)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 8),
        axis.text.x = element_text(size = 8, angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=5)) +
  geom_text(data = anatomyclass_stats, aes(x = `Anatomy class`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 2.5, color = "black", fontface = "italic")
Anatomyclass_violin

############################ INSTUTUTION CODE ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
institution_stats <- data2 %>% 
  group_by(`Institution code`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))

# We have some categories with too few points create a violin plot so need to divide the data
# Separate data based on the number of points
data_few_points <- data2 %>%
  group_by(`Institution code`) %>%
  filter(n() < 3)

data_many_points <- data2 %>%
  group_by(`Institution code`) %>%
  filter(n() >= 3)

# Create the plot
Institution_violin <- ggplot() +
  geom_violin(data = data_many_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), alpha = 0.6, lwd = 0.3) + 
  geom_dotplot(data = data_many_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), binaxis = "y", stackdir = "center", dotsize = 20, binwidth = 0.0001) + 
  geom_dotplot(data = data_few_points, aes(x = `Institution code`, y = `Endogenous(fraction)`, fill = `Institution code`), binaxis = "y", stackdir = "center", dotsize = 20, binwidth = 0.0001) +
  scale_fill_manual(name = "Endogenous(fraction)", values = rep("red", 7)) + 
  xlab("Institution code") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels = c("MCC", "MNHN", "NBC", "NHMO", "NHMUK", "RMCA", "UMB")) +
  scale_y_continuous(breaks = seq(0.0, 1.0, by = 0.2), limits = c(0.0, 1.0)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 5),
        axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(size=5)) +
  geom_text(data = institution_stats, aes(x = `Institution code`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Institution_violin

############################ REGION ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
region_stats <- data2 %>% 
  group_by(`Region`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))

# Set the order for the regions, with "Unknown" as the last category
region_order <- c("Amazon", "Caribbean and Gulf of Mexico", "Central Africa", 
                  "Eastern Indian Ocean", "South America", 
                  "West Africa", "Western Indian Ocean", "Unknown")
data2$Region <- factor(data2$Region, levels = region_order)

# Create the plot
Region_violin <- ggplot(data2, aes(x=Region, y=`Endogenous(fraction)`, fill = Region)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("yellow",8)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 20,
               binwidth = 0.0001) +
  xlab("Region") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels= region_order) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.2),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 5),
        axis.text.x = element_text(size=5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size=5)) +
  geom_text(data = region_stats, aes(x = `Region`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Region_violin

############################ YEARS B/S 1950 ############################
# First get the counts for n
# AND calculate the number of points and maximum value for each 
years_stats <- data2 %>% 
  group_by(`B/S 1950`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
Years_violin <- ggplot(data2, aes(x=`B/S 1950`, y=`Endogenous(fraction)`, fill = `B/S 1950`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("darkgreen",3)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 20,
               binwidth = 0.0001) +
  xlab("B/S 1950") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("Before","Since", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.2),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size = 8),
        axis.title.y = element_text(margin = margin(r = 5), size = 5),
        axis.text.x = element_text(size=5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5)) +
  geom_text(data = years_stats, aes(x = `B/S 1950`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Years_violin

### Let's try replotting this with the increased detail given in the Year (bins) new column...

############################ YEARS (binned) NEW ############################
# Set the desired order...
data2$`Year (bins) new` <- factor(data2$`Year (bins) new`, 
                                 levels = c( "1800-1899", "1900-1949", 
                                            "1950-1999", "Pre-1900 Unknown", 
                                            "Pre-1950 Unknown", "Unknown"))
# Now get the counts for n
# AND calculate the number of points and maximum value for each 
years_stats2 <- data2 %>% 
  group_by(`Year (bins) new`) %>% 
  summarise(n = n(), max_y = max(`Endogenous(fraction)`))
Years_violin2 <- ggplot(data2, aes(x=`Year (bins) new`, y=`Endogenous(fraction)`, fill = `Year (bins) new`)) +
  geom_violin(alpha=0.6,lwd=0.3) + 
  scale_fill_manual(name = "Endogenous(fraction)",values = rep("darkgreen",6)) +
             geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 20,
               binwidth = 0.0001) +
  xlab("Year (bins) new") +
  ylab("Endogenous DNA fraction") +
  scale_x_discrete(labels=c("1800-1899", "1900-1949", 
                            "1950-1999", "Pre-1900 Unknown", 
                            "Pre-1950 Unknown", "Unknown")) +
  scale_y_continuous(breaks = seq(0.0,1.0, by = 0.2),limits = (c(0.0,1.0))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(margin = margin(t = 10), size=8),
        axis.title.y = element_text(margin = margin(r = 5), size=5),
        axis.text.x = element_text(size=5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5)) +
  geom_text(data = years_stats2, aes(x = `Year (bins) new`, y = max_y + 0.05, label = paste0("(n = ", n, ")")),
            vjust = -0.5, size = 1.5, color = "black", fontface = "italic")
Years_violin2
```

Plot the figures - edited hereafter in Adobe Illustrator
```{r}
Sampletypesandyearsfig <- ggarrange(Sampletype_violin, Years_violin, Sampletypeclass_violin,Years_violin2,
                    labels = c("A", "C", "B", "D"),
                    ncol = 2, nrow = 2)
Sampletypesandyearsfig
ggexport(Sampletypesandyearsfig, filename = "Sampletypesandyearsfig.pdf")

instandregionanatomyfig <- ggarrange(
  Anatomyclass_violin,
  ggarrange(Institution_violin, Region_violin, ncol = 2, labels = c("F", "G"),
            widths = c(2, 3)),
  nrow = 2, 
  labels = "E",
  heights = c(3, 2)
  )
instandregionanatomyfig
ggexport(instandregionanatomyfig, filename = "instandregionanatomyfig.pdf")
```


Interpretation:
Now it seems that endogenous fraction is not vary variable across our extant 
species. Patterns remain the same across sample type variables and the institution.
When looking at region now it still seems that, although some variation may exist,
it doesn't seem significant. In looking at the temporal categories there's no obvious pattern
and maybe even the pre 1900 material looks to be better!

Test to see if there are significant differences between the time periods
```{r}
# Do t.test for Before vs Since 1950
x <- subset(data2,`B/S 1950` == "Before") 
y <- subset(data2,`B/S 1950` == "Since")
t.test(x$`Endogenous(fraction)`,y$`Endogenous(fraction)`, var.equal = FALSE)
# No significant difference - p-value = 0.1987 (Mean before is actually higher 
# than mean after)

# Do ANOVA for binned periods
one.way.aov1 <- aov(`Endogenous(fraction)`~`Year (bins) new`, data = data2)
summary(one.way.aov1)
# Filter all unknown categories out...
filter_years_data1 <- data2 %>% 
  filter(!`Year (bins) new` %in% c("Unknown", "Pre-1900 Unknown", "Pre-1950 Unknown"))
one.way.aov2 <- aov(`Endogenous(fraction)`~`Year (bins) new`, data = filter_years_data1)
summary(one.way.aov2)
```
Conclusion: Well there are significant differences when using 1) all bins but not when 
2) any bins with uncertainty are removed. Therefore there are not any discernible 
differences i.e., differences we can actually understand

Test to see if there are significant differences within the other major variables
```{r}
### Species
one.way.aov4 <- aov(`Endogenous(fraction)`~`Species`, data = data2)
summary(one.way.aov4)
# No significant differences

### Sample type
bone <- data2 %>% filter(`Sample type` == "Bone")
soft_tissue <- data2 %>% filter(`Sample type` == "Soft tissue")
t.test(bone$`Endogenous(fraction)`,soft_tissue$`Endogenous(fraction)`
       , var.equal = FALSE)

### Sample type class
# Filter unknown categories out...
filter_stc_data <- data2 %>% 
  filter(!`Sample type class` %in% c("Unknown"))
one.way.aov6 <- aov(`Endogenous(fraction)`~`Sample type class`, data = filter_stc_data)
summary(one.way.aov6)
# Significant differences!

### Institution
one.way.aov7 <- aov(`Endogenous(fraction)`~`Institution code`, data = data2)
summary(one.way.aov7)
# No significant differences

### Region
# Filter unknown out...
filter_region_data1 <- data2 %>% 
  filter(!`Region` %in% c("Unknown"))
one.way.aov8 <- aov(`Endogenous(fraction)`~`Region`, data = filter_region_data1)
summary(one.way.aov8)
# No significant differences
```

ANOVA of anatomy
```{r}
Anatomy_aov <-Anatomyplotting %>% filter(`Anatomy class` != "Unknown")
one.way.aov10 <- aov(`Endogenous(fraction)`~`Anatomy class`, data = Anatomy_aov)
summary(one.way.aov10)
# Significant differences!
```
